Quantum Hoare Type Theory

### 2020-10-30 ms-paper/ms-paper.tex
Reintroduce booleans, measure -> meas

\usepackage{listings}
\lstloadlanguages{Haskell}
\lstdefinelanguage{QHaskell}{
	language     = Haskell,
	morekeywords = {Qbit, U, C, Bit},
}

@@

Here we discuss the relevant bits from the QIO interface. QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation. The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:

@@

Complete grammar written using Ott tool at \url{https://github.com/k4rtik/QHT/blob/master/ott/qhtt.ott}.

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \type{1} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{init0} \mid \kw{init1} \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} There are primitive types for unit and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

@@

There are other propositions that involve ghosts:
\begin{itemize}
	\item uniform(\textbf{X}) is an abbreviation for $(\textbf{X}, \textbf{e}) = \frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$ where \textbf{e} is  a fresh entangled ghost variable. It means that X is uniformly distributed.
	\item separable(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{q} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable. The classical equality symbol means that the variables on both side will measure to be the exact same values. That is there is no superposition involved.
\end{itemize}

@@

\paragraph{Quantum Commands}
These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a bit).
\texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
\texttt{measure} $x$ measures the variable $x$ in the computational basis.
%\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.

\paragraph{Computations} The computations are the usual monadic return and three sequencing operations that bind their result to a variable similar to HTT. The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x =_A N$, is just syntactic sugar for let-binding.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Bit}
		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \ket{N}) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
\end{mathpar}
where x is fresh.

Initialization can only be performed for a term $N$ that can be type checked as a \type{Bit}. Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.

@@

\paragraph{Measurement}
$x \Leftarrow \kw{measure} N; E$

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
		\\ \Delta, x:\type{Bit}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bit}.Q)}
\end{mathpar}

The measurement rule can be read in a similar way as the previous rule. But in measurement, we need to additionally prove that the verification condition (written as a sequent), the precondition $P$ implies that the qubit $N$ is allocated, holds in the heap context $\Psi$. The strongest postcondition here is that the fragment of heap that only refers to location N becomes empty.

@@

In this trivial program where we initialize a new qubit with 0 and then immediately measure it, we assert that the result is equal to 0. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ proposition.

Verification steps follow:  \todo{TODO how to ensure classical equality?}

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bit} \vdash r \chk \type{Bit}}
						{q : \type{Qbit}, r : \type{Bit}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bit} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bit} . (\exists r : \type{Bit}.R)}
					\\ \cdot \vdash \kw{0} \chk \type{Bit}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init 0}; E_{23} \syn r : \type{Bit} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bit}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bit} . (\top \multimap \top \wedge Id(r, \kw{0}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bit} \{\top \wedge \kw{Id}(r, \kw{0})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {\top} r : Qbit {r =q |+\>}
      = do q <= init 0;
           q <= apply H to q;
           return q

qminus : {\top} r : Qbit {r =q |-\>}
       = do q <= init 1;
            q <= apply H to q;
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a =q (alpha |0\> + beta |1\>)}
           (a, b) : (Qbit o* Qbit)
        {(a, b) =q (alpha |00\> + beta |11\>)}
      = \a.do b <= init 0;
              (a, b) <= apply CNot to [a, b];
              return (a, b)
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\todo{TODO: how is classical equality ensured?}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
         = do (qa, qb) <- bell00;
              a <= measure qa;
              b <= measure qb;
              return (a, b)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi q : Qbit. \Pi a : Qbit
        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- a is entangled with ghost var e
           (m1, m2) : (Bit o* Bit)
        {  (m1, m2) = 00 /\ e =q (alpha |0\> + beta |1\>)
         \/ (m1, m2) = 01 /\ e =q (alpha |1\> + beta |0\>)
         \/ (m1, m2) = 10 /\ e =q (alpha |0\> - beta |1\>)
         \/ (m1, m2) = 11 /\ e =q (alpha |1\> - beta |0\>)}
      = \q.\a.do (q, a) <= apply CNOT to [q, a];
                 q <= apply H to q;
                 rq <= measure q;
                 ra <= measure a;
                 return (rq, ra)
\end{lstlisting}

\subsection{Bob's circuit}
\todo{TODO provide correct propositions using ghost e}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bit. \Pi b : Qbit
      {(b, e) =q |\Phi+\>}
         r : Qbit
      {\top}
      = \m1.\m2.\b.do (m1, b) <= apply C_Z to [m1, b];
                      (m2, b) <= apply C_X to [m2, b];
                      return b
\end{lstlisting}

>>>>

\usepackage{listings}
\lstloadlanguages{Haskell}
\lstdefinelanguage{QHaskell}{
	language     = Haskell,
	morekeywords = {Qbit, U, C, meas, apply, to},
}

@@

Here we discuss the relevant concepts from the QIO interface. QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation. The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:

@@

%Complete grammar written using Ott tool at \url{https://github.com/k4rtik/QHT/blob/master/ott/qhtt.ott}.

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} There are primitive types for unit, booleans, and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

@@

There are other propositions that involve ghosts:
\begin{itemize}
	\item uniform(\textbf{X}) is an abbreviation for $(\textbf{X}, \textbf{e}) = \frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$ where \textbf{e} is  a fresh entangled ghost variable. It means that X is uniformly distributed.
	\item separable(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{q} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable. The classical equality symbol means that the variables on both side will measure to be the exact same value. That is, there is no superposition involved.
\end{itemize}

In addition to these quantum-specific propositions, the system also supports first-order logic (not shown) for reasoning about the pure fragment of the language.

@@

\paragraph{Quantum Commands}
These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically \ket{0} or \ket{1} if N is a boolean).
\texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
\texttt{meas} $x$ measures the variable $x$ in the computational basis.
%\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.

\paragraph{Computations} The computations are the usual monadic return and three sequencing operations that bind their result to a variable similar to HTT. The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x = N$, is just syntactic sugar for let-binding.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Bool}
		\\ \Delta, x : \type{Qbit}; P \circ (x =_q \mathit{bool2ket}(N)) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
\end{mathpar}
where x is fresh.

Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}. Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ is equal to the \ket{0} or \ket{1}; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.

@@

\paragraph{Measurement}
$x \Leftarrow \kw{meas} N; E$

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N =_q -)
		\\ \Delta, x:\type{Bool}; P \circ ((N =_q -) \multimap \top) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{meas} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
\end{mathpar}

The measurement rule can be read in a similar way as the previous rule. But in measurement, we need to additionally prove that the verification condition (written as a sequent), the precondition $P$ implies that the qubit $N$ is allocated, holds in the heap context $\Psi$. The strongest postcondition here is that the fragment of heap that only refers to location N becomes empty. \todo{TODO: need to introduce ghost variable for measurement}

@@

In this trivial program where we initialize a new qubit with false and then immediately measure it, we assert that the result is false. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ proposition.

Verification steps follow:

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{meas} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{0} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{0}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{0})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {\top} r : Qbit {r =q |+\>}
      = do q <= init false;
           q <= apply H to q;
           return q

qminus : {\top} r : Qbit {r =q |-\>}
       = do q <= init true;
            q <= apply H to q;
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a =q (alpha |0\> + beta |1\>)}
           (a, b) : (Qbit o* Qbit)
        {(a, b) =q (alpha |00\> + beta |11\>)}
      = \a.do b <= init false;
              (a, b) <= apply CNot to [a, b];
              return (a, b)
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\todo{TODO: how is classical equality ensured?}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
         = do (qa, qb) <- bell00;
              a <= meas qa;
              b <= meas qb;
              return (a, b)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi q : Qbit. \Pi a : Qbit
        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- a is entangled with ghost var e
           (m1, m2) : (Bool o* Bool)
        {  (m1, m2) = 00 /\ e =q (alpha |0\> + beta |1\>)
         \/ (m1, m2) = 01 /\ e =q (alpha |1\> + beta |0\>)
         \/ (m1, m2) = 10 /\ e =q (alpha |0\> - beta |1\>)
         \/ (m1, m2) = 11 /\ e =q (alpha |1\> - beta |0\>)}
      = \q.\a.do (q, a) <= apply CNOT to [q, a];
                 q <= apply H to q;
                 rq <= meas q;
                 ra <= meas a;
                 return (rq, ra)
\end{lstlisting}

\subsection{Bob's circuit}
\todo{TODO provide correct propositions using ghost e}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bool. \Pi b : Qbit
      {(b, e) =q |\Phi+\>}
         r : Qbit
      {\top}
      = \m1.\m2.\b.do (m1, b) <= apply C_Z to [m1, b];
                      (m2, b) <= apply C_X to [m2, b];
                      return b
\end{lstlisting}
